A Formalization of Complete Discrete Valuation Rings and Local Fields

### 2023-12-11 HAL/CPP_paper_v2.tex
comments by fil

\section{Introduction}\label{sec:intro}
The vague idea that geometric intuition and algebraic rigor can fruitfully interact is an old theme, certainly predating the modern approach by Descartes and subsequently by Newton and Leibniz. 
In contemporary commutative algebra, and especially
after the advent of scheme theory by Grothendieck around 1960, this connection has become even tighter: geometric concepts and techniques are often borrowed for a wide range of applications. In this work we are concerned with the concept of \emph{local field}, a fundamental notion in algebraic number theory whose origin, and still many applications, comes from geometry and diophantine questions. Suppose, for instance, that one is interested in the integral solutions $(a_1,a_2,a_3)$ to $X_1^2+X_2^2+X_3^2=0$: clearly, since for positivity reasons the only \emph{real} solution is $(0,0,0)$, there cannot be any other integral solutions. On the other hand, $X_1^2+ X_2^2-3X_3^2=0$ certainly has real solutions, yet has no non-trivial solutions in $\FF[3]$, the field with $3$ elements. Now, an integral solution $(a_1,a_2,a_3)$ with $a_1\equiv a_2\equiv 0\pmod{3}$ must be $(0,0,0)$, as can be seen by comparing the occurrences of $3$ in the factorizations of $a_1^2+a_2^2$ and of $3a_3^2$.

From a geometric perspective, one can regard the previous as a kind of ``local analysis'': interpreting the primes as the points of a geometric object attached to $\ZZ$, if a ``global'' solution $(a_1,a_2,a_3)\in\ZZ^3$ exists, we could obtain infinitely many solutions $(\overline{a_1},\overline{a_2},\overline{a_3})\in\FF^3$ ``around each point'', for varying $p$, as well as a solution $((a_1 : \RR),(a_2 : \RR),(a_3 : \RR))\in\RR^3$ over the reals. If at some point no solution can be found, then no global solution exists. Yet, in this analogy, the fields $\FF$ and $\RR$ are very different: the former is finite, with trivial discrete geometry and of positive characteristic, while the latter has a rich metric structure and characteristic $0$. Now, for every $p$, one can rather consider the field of $p$-adic numbers $\QQ_p$ which is, in this perspective, a better analogue for $\RR$ than the several $\FF$: it has a metric with respect to which it is complete (Cauchy sequences converge), and it has characteristic $0$ (so, in particular, it contains $\QQ$). To define it, one observes that for every prime $p$ there exists an absolute value $\lvert\cdot\rvert_p$ on $\QQ$ for which numbers that are highly divisible by $p$ are close to $0$. Having an absolute value, it is possible to speak about convergence and about Cauchy sequences: exactly as for the euclidean absolute value, one can find Cauchy sequences that do not converge to any rational, and $\QQ_p$ is defined as the \emph{completion} of $\QQ$ with respect to~$\lvert\cdot\rvert_p$. It is the ``smallest'' field where all $\lvert\cdot\rvert_p$-Cauchy sequences converge. Crucially, it still bears strong connections with the prime $p$: for instance, if a monic polynomial in $\ZZ[X]$ has a simple root in $\FF$, then it also has one in $\QQ_p$.

The local analysis described before is an example of the ``local-to-global'' principle: to solve a problem in $\ZZ$, or $\QQ$, one can try to solve it in $\QQ_p$, for all $p$, and in $\RR$. For a notorious example, it has long been known that the so-called ``first case'' of Fermat's Last Theorem can be solved using class field theory (see~\cite{LenSte97}), which is a deep theory where the global theorems are obtained by a beautiful patching of the local ones: we refer to \cite{Tat67}~for an excellent account of this example. It should also be emphasized that ultimately Wiles' very proof of Fermat's Last Theorem makes heavy use of the local-to-global principle in the framework of Galois representations: we refer to~\cite{Fre09} for a summary of this technique.

@@

We also provide a bundled version of this definition, called \code{uniformizer}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/basic.lean#L144}{\extlink}. Since any uniformizer is necessarily a member of the unit ball, we decided to make the field $\pi$ in the definition below a term of type \code{vR.integer} (as opposed to type \code{R}). The bundled definition is more convenient to prove results that involve fixing a uniformizer.
\begin{lstlisting}
structure uniformizer :=
($\pi$ : vR.integer)
(valuation_eq_neg_one : is_uniformizer vR $\pi$)
\end{lstlisting}

\paragraph{Extensions and \code{valued} instances}
In \S\ref{subsec:extensions} we construct, for each field $K$ complete with respect to a discrete valuation~$v$ and every finite extension $L/K$, a unique valuation $v_L$ on $L$, recorded as \code{(v_L : valuation L}$\;\wzmZZ)$. However, at this level of generality we do not put a global \code{valued} instance on $L$: doing so would create infinitely many ``diamonds'', by which we mean the existence of two different procedures to endow an object with a certain mathematical structure (represented by two terms of the given structure that are propositionally, but not definitionally\mi{, equal}). This leads to problems when exploiting the full force of the type-class inference mechanism: indeed, suppose that in a given context $\Gamma$, a term $t:T$ is well-typed assuming the existence of a term \code{s:S} for a certain structure \code{S}. Given two terms \code{s}$_1$,\code{s}$_2$\code{:S} that are propositionally equal, the corresponding terms \code{t}$_1$,\code{t}$_2$\code{:T} are different, yet carry the same mathematical information. Now, if the type-class inference mechanism can produce both terms \code{s}$_1$ and \code{s}$_2$ --- say, inside a proof --- the user would face the problem of having two different terms that look indistinguishable, which clearly leads to unexpected problems. In our setting, for every field $K$ that is complete with respect to a discrete valuation $v$, the original valuation is propositionally but not definitionally equal to its trivial extension $v_K$, resulting in two terms of the structure of a \code{valued} field on $K$: to avoid any diamond, we need to exclude the second from the inference search, and therefore we need to avoid declaring a structure of \code{valued} field on the trivial extension $K$ of $K$: this forces us to avoid declaring such a structure on \emph{every} extension $L/K$. Nevertheless, we provide a lemma \code{trivial_extension_eq_valuation}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/trivial_extension.lean#L58}{\extlink} proving the equality $v=v_K$.

@@

\paragraph{The \code{normed_field} and the \code{valued} instances}
Recall from~\S\ref{subsec:extensions} that to each discrete valuation $v$ on a field $K$ we can associate a nonarchimedean multiplicative norm $\lvert\cdot\rvert_K$. If $v$ is the unique discrete valuation on $K$, we often register it as a \code{valued} instance:
\begin{lstlisting}
    {K : Type*} [field K] [hv : valued K $\wzmZZ$] 
\end{lstlisting}
In this situation, $\lvert\cdot\rvert_K$ would be the unique nontrivial norm on $K$, up to rescaling, and we would like to record a corresponding \code{normed_field} instance on $K$:
\begin{lstlisting}
    {K : Type*} [normed_field K $\wzmZZ$] 
\end{lstlisting}
Note that the datum of a field is embedded into the definition of the \code{normed_field} class, while the class \code{valued} takes the field structure as an argument. Therefore, declaring a \code{normed_field} instance on every field that carries a \code{valued} one leads to a loop in the typeclass inference system\footnote{\lean[4] can detect these kinds of simple loops, so this should not be an issue once our project has been ported.}.

\mi{We give a concrete example to illustrate this problem, for which we studied
a trace of the instance search using the option} \code{trace.class_instances} \mi{ to track the typeclass inference process.

Suppose that we had declared} \code{discretely_normed_field} \mi{as an instance instead of as a definition, so that every discretely valued field}
\begin{lstlisting}
(K : Type*) [field K] [hv : valued K $\wzmZZ$]
  [is_discrete hv.v]
\end{lstlisting}
\mi{would inherit a} \code{normed_field} \mi{instance.
Under this hypothesis, the typeclass inference system would get into an infinite loop while trying to find an instance of
}
\code{
valuation_ring hv.v.valuation_subring.to_subring.}

@@

However, \mi{we remark that} no problem arises when putting both a \code{valued} and a \code{normed_field} instance on a specific field, like $\QQ_p$.

@@

\begin{acks}
We would like to thank Riccardo Brasca, Mario Carneiro, Antoine Chambert-Loir and Rob Lewis for interesting discussions about possible formalizations of the $p$-adic numbers, Heather Macbeth for helpful conversations about the formalization of Laurent series, and Yaël Dillies for providing access to some results about well-founded relations. We also thank the \mathlib community as a whole for their support during the completion of this project.

>>>>

\section{Introduction}\label{sec:intro}
The vague idea that geometric intuition and algebraic rigor can fruitfully interact is an old theme, certainly predating the modern approach by Descartes and subsequently by Newton and Leibniz. 
In contemporary commutative algebra, and especially
after the advent of scheme theory by Grothendieck around 1960, this connection has become even tighter: geometric concepts and techniques are often borrowed for a wide range of applications. In this work we are concerned with the concept of \emph{local field}, a fundamental notion in algebraic number theory whose origin, and still many applications, comes from geometry and diophantine questions. Suppose, for instance, that one is interested in the integral solutions $(a_1,a_2,a_3)$ to $X_1^2+X_2^2+X_3^2=0$: clearly, since for positivity reasons the only \emph{real} solution is $(0,0,0)$, there cannot be any other integral solutions. On the other hand, $X_1^2+ X_2^2-3X_3^2=0$ certainly has real solutions, yet has no non-trivial solutions in $\FF[3]$, the field with $3$ elements, as can be seen by analyzing the finitely many possibilities. Hence, if $(a_1,a_2,a_3)$ is an integral solution to this equation, its reduction modulo 3 must be the trivial solution in $\FF[3]$, so that $a_1, a_2$ and $a_3$ are all divisible by $3$. This implies $a_1 = a_2 = a_3 = 0$, since otherwise we would have an equality $a_1^2 + a_2^2 = 3a_3^2$, in which the highest power of $3$ dividing the left hand side is even, while the highest power of $3$ dividing the left hand side is odd, yielding a contradiction.

From a geometric perspective, one can regard the previous as a kind of ``local analysis'': interpreting the primes as the points of a geometric object attached to $\ZZ$, if a ``global'' solution $(a_1,a_2,a_3)\in\ZZ^3$ exists, then for each prime $p$, we obtain a solution $(\overline{a_1},\overline{a_2},\overline{a_3})\in\FF^3$ by reducing the coordinates modulo $p$ --- we think of this as a local solution ``around the point $p$'' --- as well as a solution  $((a_1 : \RR),(a_2 : \RR),(a_3 : \RR))\in\RR^3$ over the reals. \mi{Conversely}, if one of these local solutions fails to exists, this means that the equation cannot have a global solution. Yet, in this analogy, the fields $\FF$ and $\RR$ are very different: the former is finite, with trivial discrete geometry and of positive characteristic, while the latter has a rich metric structure and characteristic $0$. Now, for every $p$, one can rather consider the field of $p$-adic numbers $\QQ_p$ which is, in this perspective, a better analogue for $\RR$ than the several $\FF$: it has a metric with respect to which it is complete (Cauchy sequences converge), and it has characteristic $0$ (so, in particular, it contains $\QQ$). To define it, one observes that for every prime $p$ there exists an absolute value $\lvert\cdot\rvert_p$ on $\QQ$ for which numbers that are highly divisible by $p$ are close to $0$. Having an absolute value, it is possible to speak about convergence and about Cauchy sequences: exactly as for the euclidean absolute value, one can find Cauchy sequences that do not converge to any rational, and $\QQ_p$ is defined as the \emph{completion} of $\QQ$ with respect to~$\lvert\cdot\rvert_p$. It is the ``smallest'' field where all $\lvert\cdot\rvert_p$-Cauchy sequences converge. Crucially, it still bears strong connections with the prime $p$: for instance, if a monic polynomial in $\ZZ[X]$ has a simple root in $\FF$, then it also has one in $\QQ_p$.

The local analysis described before is an example of the ``local-to-global'' principle: to solve a problem in $\ZZ$, or $\QQ$, one can try to solve it in $\QQ_p$, for all $p$, and in $\RR$. A famous application of the local-to-global principle is \mi{found in} the Hasse–Minkowski Theorem, which states that if $Q(x)$ is a quadratic form with rational coefficients and $c \in \unit{\QQ}$, then the equation $Q(x) =c$ is solvable over $\QQ$ if and only if it is solvable over $\RR$ and over every $\QQ_p$.

For another notorious example, it has long been known that the so-called ``first case'' of Fermat's Last Theorem can be solved using class field theory (see~\cite{LenSte97}), which is a deep theory where the global theorems are obtained by a beautiful patching of the local ones: we refer to \cite{Tat67}~for an excellent account of this example. It should also be emphasized that ultimately Wiles' very proof of Fermat's Last Theorem makes heavy use of the local-to-global principle in the framework of Galois representations: we refer to~\cite{Fre09} for a summary of this technique.

@@

We also provide a bundled version of this definition, called \code{uniformizer}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/basic.lean#L144}{\extlink}. \mi{By ``bundled'' we mean that, in this version, the element $\pi$ is recorded as a field of the structure} \code{uniformizer}, \mi{instead of being a parameter as in the previous definition} \code{is_uniformizer}.

Since any uniformizer is necessarily a member of the unit ball, we decided to make the field $\pi$ in the definition below a term of type \code{vR.integer} (as opposed to type \code{R}). The bundled definition is more convenient to prove results that involve fixing a uniformizer.
\begin{lstlisting}
structure uniformizer :=
($\pi$ : vR.integer)
(valuation_eq_neg_one : is_uniformizer vR $\pi$)
\end{lstlisting}

For an in-depth discussion of bundled versus unbundled representations \fil{in Lean and in Coq}, we refer the reader to \cite{Baa22} or \cite{typeclassesCoq}.

\paragraph{Extensions and \code{valued} instances}
In \S\ref{subsec:extensions} we construct, for each field $K$ complete with respect to a discrete valuation~$v$ and every finite extension $L/K$, a unique valuation $v_L$ on $L$, recorded as \code{(v_L : valuation L}$\;\wzmZZ)$. However, at this level of generality we do not put a global \code{valued} instance on $L$: doing so would create infinitely many ``diamonds'', by which we mean the existence of two different procedures to endow an object with a certain mathematical structure (represented by two terms of the given structure that are propositionally, but not definitionally, equal). This leads to problems when exploiting the full force of the type-class inference mechanism: indeed, suppose that in a given context $\Gamma$, a term $t:T$ is well-typed assuming the existence of a term \code{s:S} for a certain structure \code{S}. Given two terms \code{s}$_1$,\code{s}$_2$\code{:S} that are propositionally equal, the corresponding terms \code{t}$_1$,\code{t}$_2$\code{:T} are different, yet carry the same mathematical information. Now, if the type-class inference mechanism can produce both terms \code{s}$_1$ and \code{s}$_2$ --- say, inside a proof --- the user would face the problem of having two different terms that look indistinguishable, which clearly leads to unexpected problems. In our setting, for every field $K$ that is complete with respect to a discrete valuation $v$, the original valuation is propositionally but not definitionally equal to its trivial extension $v_K$, resulting in two terms of the structure of a \code{valued} field on $K$: to avoid any diamond, we need to exclude the second from the inference search, and therefore we need to avoid declaring a structure of \code{valued} field on the trivial extension $K$ of $K$: this forces us to avoid declaring such a structure on \emph{every} extension $L/K$. Nevertheless, we provide a lemma \code{trivial_extension_eq_valuation}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/trivial_extension.lean#L58}{\extlink} proving the equality $v=v_K$.

@@

\paragraph{The \code{normed_field} and the \code{valued} instances}
Recall from~\S\ref{subsec:extensions} that to each discrete valuation $v$ on a field $K$ we can associate a nonarchimedean multiplicative norm $\lvert\cdot\rvert_K$. If $v$ is the \fil{unique} discrete valuation on $K$, we often register it as a \code{valued} instance:
\begin{lstlisting}
    {K : Type*} [field K] [hv : valued K $\wzmZZ$] 
\end{lstlisting}
In this situation, $\lvert\cdot\rvert_K$ would be the \fil{unique} nontrivial norm on $K$, up to rescaling, and we would like to record a corresponding \code{normed_field} instance on $K$:
\begin{lstlisting}
    {K : Type*} [normed_field K $\wzmZZ$] 
\end{lstlisting}
Note that the datum of a field is embedded into the definition of the \code{normed_field} class, while the class \code{valued} takes the field structure as an argument. \fil{\st{Therefore,}It follows that} declaring a \code{normed_field} instance on every field that carries a \code{valued} one leads to a loop in the typeclass inference system\footnote{\lean[4] can detect these kinds of simple loops, so this should not be an issue once our project has been ported \fil{to \lean[4]}.}, \fil{as can be seen in the following example:\st{We give a concrete example to illustrate this problem, for which we studied
a trace of the instance search using the option} \code{trace.class_instances} \st{to track the typeclass inference process.}}

Suppose that we had declared \code{discretely_normed_field} as an instance instead of as a definition, so that every discretely valued field
\begin{lstlisting}
(K : Type*) [field K] [hv : valued K $\wzmZZ$]
  [is_discrete hv.v]
\end{lstlisting}
\mi{would inherit a} \code{normed_field} \mi{instance.
Under this hypothesis, the typeclass inference system would get into an infinite loop while trying to find an instance of
}
\code{
valuation_ring hv.v.valuation_subring.to_subring.}

@@

However, \fil{\st{we remark that}observe that} no problem arises when putting both a \code{valued} and a \code{normed_field} instance on a specific field, like $\QQ_p$\fil{, because\dots}

@@

\begin{acks}
We would like to thank Riccardo Brasca, Mario Carneiro, Antoine Chambert-Loir and Rob Lewis for interesting discussions about possible formalizations of the $p$-adic numbers, Heather Macbeth for helpful conversations about the formalization of Laurent series, Yaël Dillies for providing access to some results about well-founded relations\ \fil{and Cyril Cohen for answering some questions about $p$-adic numbers in Coq}. We also thank the \mathlib community as a whole for their support during the completion of this project.
